# Working with DBMS using R



**Introduction**

- Various DBMSs can be worked with using R
- The specific way that this is done differ between the DBMSs and operating system





**Working with ODBC-compliant DBMS with R**
- **Some terms**
  - **DBMS**
    - A system for managing data 
    - Examples include Microsoft SQL Server, Microsoft Access, Oracle, MySQL, PostgreSQL, MariaDB, SQLite
  - **Open Database Connectivity (ODBC)**
    - An ODBC is a connection between a software/application (e.g. R, Python, SPSS, Excel) and a DBMS through an ODBC interface/system (a standardised connection)
  - **ODBC-compliant DBMS**
    - ODBC-compliant DBMSs are DBMSs that adheres to the ODBC standards to allow applications to access them through ODBC (a standardised way of doing so)
    - ODBC-compliant DBMSs include Microsoft SQL Server, Microsoft Access, MySQL, PostgreSQL, Oracle Database, IBM Db2, SQLite, MariaDB, Amazon Redshift, Google BigQuery, etc.
- **The Process**
  - In general, working with an ODBC-compliant DBMS from an application (e.g. R and RStudio) necessarily involves, in addition to the application and the DBMS, an ODBC interface, an ODBC manager, and a DBMS-specific ODBC driver for mediating the connection between the application and the DBMS. The ODBC interface is usually associated with the application, specifically, it is either a functionality of the application or an extension to the application (e.g. packages/modules/plug-ins). The ODBC interface translates application-specific request calls from the application to an ODBC call, which is a call adhering to the ODBC standards, through its ODBC functions, it then establishes a connection with and sends the ODBC call to the DBMS-specific ODBC driver either directly or indirectly through the ODBC manager, which manages one or more DBMS-specific ODBC drivers. The DBMS-specific ODBC driver translates the ODBC call to a DBMS-specific call (which may involve DBMS-specific SQL), it then establishes a connection to and sends the DBMS-specific call to the DBMS (e.g. its server/engine) where the call is executed. If the call involves extracting information from the DBMS, the information is sent from the DBMS to its DBMS-specific ODBC driver, the DBMS-specific ODBC driver translates it to an ODBC information, finally, the ODBC interface fetches the ODBC information from the DBMS-specific driver and may further process it (e.g. translation) if there is an upper layer (e.g. application). 

  
  
```{r}
?DBI::dbFetch()
```

  
  **Open Database Connectivity (ODBC)**

- **Concept**
  - An ODBC is a connection between an software/application (e.g. R, Python, SPSS, Excel) and a DBMS
  
**ODBC interface**

- **Concept**
  - An API that manages one or more ODBCs
  - Provides a standard and general set of calls/commands (via ODBC functions) for softwares/applications to ODBC compliant DBMS (e.g. executing an SQL script), this standarisation and generalisation enables an ODBC to be independent of applications
  - Developed by Microsoft in early 1990s
  
  
**ODBC driver**

- **Concept**
  - An ODBC driver provides a connection between an ODBC interface and a DBMS
  - An ODBC interface connects to different ODBC drivers as different DBMSs operate differently thus have their own specific DBMS function calls
  - An ODBC driver translates ODBC function calls into DBMS-specific function calls and passes them to the specific DBMS
  - This enables an ODBC interface to be independent of DBMSs
  - E.g. The old SQL Server and the new ODBC for SQL Server (these are drivers, despite having the word "servers" in their names)


```{r}

```


**DSN**
- A parametric data structure/format that contains information for an ODBC driver to connect to a specific part of a DBMS (e.g. server/engine/database/table)
- DSNs are ODBC driver-specific, but they tend to be simple and short with common parameters 
- Parameters of the DSN for an ODBC driver includes the name of the engine, name of the server, name of the database, name of the table, username and password for access


**Windows**
- **In general**
  - 
- ODBC Interface (API)
- ODBC Drivers

- 
- ODBC Data Source Administrator
- ODBC Manager 
- ODBC driver 




- 
- Set up a connection between the ODBC interface and the database 
- ODBC driver (install)
- The ODBC Data Source Administrator (64-bit)

- Set up a connection between R and the ODBC interface 
- R

```{r}
DBI::db
```

